---
title: "Final project"
author: "Herong Wang"
format: 
  html:
    embed-resources: true
    code-fold: true
    code-summary: "Show the code"
editor: visual
---

## Basic set up: load packages and read in data

```{r}
# load libraries
library(pacman)
p_load(tidyverse, haven, interactions)

preg <- read_dta("C:/Users/herongw/Desktop/Umich_Phd/2024fall/STAT506/Final_project/data/2017_2019_Preg.dta")

female <- read_dta("C:/Users/herongw/Desktop/Umich_Phd/2024fall/STAT506/Final_project/data/2017_2019_Femresp.dta")

male <- read_dta("C:/Users/herongw/Desktop/Umich_Phd/2024fall/STAT506/Final_project/data/2017_2019_MaleData.dta")

```

## Data cleaning

#### Outcome: PREGEND1 from Female Pregnancy dataset.
```{r}
# original variable PREGEND1 will be collapsed into two levels. Miscarriage (1), stillbirth (2), abortion (3), ectopic or tubal pregnancy (4) will be all coded as 1 while live birth by Cesarean/vaginal delivery (5, 6) will be coded as 0. Refused (8) or Don’t know (9) responses will be coded as missing (NA)

preg.fin <- preg %>% 
  filter(pregordr == 1) %>% 
  select(caseid, PREGEND1) %>% 
  mutate(pregend.clean = case_when(PREGEND1 %in% c(5,6) ~ 0,
                                   PREGEND1 %in% c(1,2,3,4) ~ 1,
                                   TRUE ~ NA))

table(preg.fin$PREGEND1, preg.fin$pregend.clean, useNA = "always")

preg.fin$pregend.clean <- factor(preg.fin$pregend.clean, levels = c(0,1), labels = c("Live birth", "Adverse birth"))
```

#### Exposure: mother’s education level. HIGRADE from Female Respondent dataset 
```{r}
education.m <- female %>% 
  select(caseid, higrade)
```

#### Covariates: mother’s age at the time of conception (AGECON), father’s age at pregnancy’s end date/live birth date (HPAGEEND/ HPAGELB)
```{r}
cor <- preg %>% 
  filter(pregordr == 1) %>% 
  select(caseid, agecon, hpageend, hpagelb) %>% 
  mutate(father.age = case_when(!is.na(hpageend)&is.na(hpagelb) ~ hpageend,
                                is.na(hpageend)&!is.na(hpagelb) ~ hpagelb,
                                TRUE ~ NA)) %>% 
  mutate(father.age = if_else(father.age %in% c(98,99), NA, father.age))

```

#### Interaction variable: Father’s education level CHEDMARN frm Female response data
```{r}
education.f <- female %>% 
  select(caseid, chedmarn) %>% 
  mutate(father.education = case_when(chedmarn %in% c(1,2,3,4,5) ~ "College or less",
                                      chedmarn == 6 ~ "More than college",
                                      TRUE ~ NA))

education.f$father.education <- factor(education.f$father.education, levels = c("High school or less", "College or less", "More than college"))
table(education.f$chedmarn, education.f$father.education, useNA = "always")
```

#### Merge datasets
```{r}
dat <- preg.fin %>% 
  inner_join(education.m, by = "caseid") %>% 
  inner_join(cor, by = "caseid") %>% 
  inner_join(education.f, by = "caseid")

## only include non-missing observations
var <- c("pregend.clean", "higrade", "agecon", "father.age",  "father.education")

fi.dat <- dat %>% 
  select(caseid, all_of(var)) %>% 
  filter(complete.cases(.))
```

#### fit model
```{r}

## crude model without covariates and interactive variable
logit.crude <- glm(pregend.clean ~ higrade, data = fi.dat, family = "binomial")
summary(logit.crude)
exp(cbind(OR = coef(logit.crude), confint(logit.crude)))

## adjusted model with covariates but no interaction
logit.adj <- glm(pregend.clean ~ higrade + agecon + father.age, data = fi.dat, family = "binomial")
summary(logit.adj)
exp(cbind(OR = coef(logit.adj), confint(logit.adj)))

## interactive model with both covariates and interaction
logit.int <- glm(pregend.clean ~ higrade + agecon + father.age + father.education + higrade*father.education, data = fi.dat, family = "binomial")
summary(logit.int)
exp(cbind(OR = coef(logit.int), confint(logit.int)))
```

#### visualization
```{r}
interactions::cat_plot(logit.int, pred = higrade, modx = father.education, interval = TRUE, vary.lty = TRUE, geom = "line",
                       x.label = "Mother's education years", y.label = "Probabilities of ending pregnancy adversely", 
                       main.title = "The association of mother's education with adverse pregnancy, differed by father's education",
                       legend.main = "Father's education")



```

